### 2025-11-11 cluster-analysis.qmd
20251111 - cluster analysis models and fit criteria

@@

@@

We can perform a cluster analysis using the `mclust::mclustBIC()`, `mclust::mclustICL()`, and `mclust::mclustBootstrapLRT()` functions of the `mclust` package [@R-mclust; @Scrucca2023_packages].

>>>>

There are several model fit criteria we can use to identify the optimal number of tiers in terms of the tradeoff between fit and parsimony.
For the Bayesian Information Criterion (BIC), closer to zero is better.
For the Integrated Complete-data Likelihood (ICL), higher is better.
For the bootstrapped likelihood ratio tests, a significant likelihood ratio test indicates that the model with more clusters fits significantly better than the model with fewer clusters.

For univariate cluster analyses in the `mclust` package [@R-mclust; @Scrucca2023_packages], there are two types of models: ones with equal variance across clusters (`modelName = "E"`) and ones with different variances across clusters (`modelName = "V"`).
Allowing different variances across clusters is a more complex model than forcing the clusters to have equal variances.

For multivariate cluster analyses in the `mclust` package [@R-mclust; @Scrucca2023_packages], models are specified by a combination of volume, shape, and orientation.
The model name is specified such that the first letter corresponds to the volume, the second letter corresponds to shape, and the third letter corresponds to orientation.
In terms of volume, models can be of equal volume (`E`) or variable volume (`V`) across clusters.
In terms of shape, models can be of equal (`E`), variable (`V`), or identity (`I`) shape.
In terms of orientation, models can be of equal (`E`), variable (`V`), or identity (`I`) orientation.
The varying types of volume, shape, and orientation combine to produce three general types of covariance structures: spherical (low complexity), diagonal (moderate complexity), and ellipsoidal (high complexity).
Models that allow greater differences across clusters (i.e., more `V`'s) are more flexible and typically more complex.

@@

We can evaluate how many clusters to keep, based on the tradeoff between fit and parsimony, using the `mclust::mclustBIC()` (BIC), `mclust::mclustICL()` (ICL), and `mclust::mclustBootstrapLRT()` (bootstrapped likelihood ratio tests) functions of the `mclust` package [@R-mclust; @Scrucca2023_packages].

@@
